Helixer: Cross-species Gene Annotation Of Large Eucaryotic Genomes Using Deep Learning

### 2020-05-14 paper/main.tex
fin brain dump, this probably got a little long...

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{alisandra.denton@hhu.de}{alisandra.denton@hhu.de}\\  
% normally contact is last author / person with permanent position (i.e. Andreas). 
% Not sure how important that is, but if we're skipping that, then we can change it back to Felix
\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.}

@@

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences \ref{fig:examples} % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d).

@@

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes \ref{}. % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron \ref{}, there was more coverage when 
Helixer said CDS. Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron \ref{}. 
These patterns were consistent in direction but varied in magnitude in the individual species. % double check
RNAseq based comparison of Helixer and the reference was less clear cut.
-Helixer is good at differentiating ig, ref better at introns
-diffs between species. 
-Helixer is substantially better for Tcacao, with a "poor Sub Genic F1"
-taken together: competitive quality predictions, in some cases the mistakes really are in the reference / we wouldn't want 100\% accuracy against the reference

\section{Discussion}
We propose a novel way to develop gene calling models that are able to perform 
cross-species genomic annotiation for a wide variety of large eucaryotic genomes. 
Helixer does this while outperforming AUGUSTUS on a base pair wise level and 
consistency with independent RNASeq data.

Eliminating the need for a close genome annotation being present for pretraining.

We found our models to be highly sensitive to the training genomes we chose. A 
different set could lead to a significant shift in strengths and weaknesses of the 
model and a larger and more spread out set of high quality genomes could also result 
in a wider range of genomes with decent predictions. For this, more computational 
resources would be required.

@@

@@

>>>>

\abstract{\textbf{Motivation:} Current state-of-the-art tools for the \textit{de novo} annotation of 
genes in eukaryotic genomes have to be specifically fitted for each species and still often 
produce annotations that can be improved much further. The fundamental algorithmic 
architecture for these tools has remained largely unchanged for about two decades, limiting
learning capabilities. Here, we set out to improve the cross-species annotation of
genes from DNA sequence alone with the help of deep learning. The goal is to eliminate 
the dependency on a closely related gene model while also improving the predictive
quality in general with a fundamentally new architecture.\\
\textbf{Results:} We present Helixer, a framework for the development and usage of 
a cross-species deep learning model that improves significantly on performance and
generalizability when compared to more traditional methods. We evaluate our 
approach by building a single model for the annotation of 186 animal genomes as 
well as one for 51 plants. Our predictions are shown to be much less sensitive to
the length of the genome than those of a current state-of-the-art tool. We also 
present two novel post-processing techniques that each worked to further strengthen 
our annotations and show in-depth results of an RNA-Seq based comparison of our 
predictions.\\
\textbf{Availability:} The source code of this work is available at 
\url{https://github.com/weberlab-hhu/HelixerPrep} [change] under the GNU General 
Public License v3.0.\\
\textbf{Contact:} \href{andreas.weber@hhu.de}{andreas.weber@hhu.de}\\

\textbf{Supplementary information:} Supplementary data are available at 
\textit{Bioinformatics} online.}

@@

As an independent evaluation and to get a better qualitative understanding of
performance we compared Helixer predictions, Augustus predictions and the
reference to RNAseq coverage data. RNAseq was processed for three each of plants
and animals chosen because they had varied performance quality relative to Augustus. 
RNAseq coverage provides support for an 
exonic annotation (UTR or CDS), spliced coverages provides support for an 
annotation of intron, and neither coverage nor spliced coverage is expected 
for intergenic annotations. Looking at selected subsequences \ref{fig:examples} % and supplemental figures.
we identified cases where RNAseq supported both the annotations of the reference 
and Helixer (a), neither (b), the reference, but not Helixer (c),
and Helixer, but not the reference (d).

@@

Coverage and Spliced coverage were broken down by the confusion matrix (of both
Helixer vs Augustus and Helixer vs the reference) for all genomes \ref{}. % H vs A main, H vs R supp
Where tools were in agreement (far left), coverage and spliced coverage closely matched
expectations. Specifically for CDS:CDS and UTR:UTR most base pairs had some, and many
had moderate or high coverage. The same pattern was seen for intron:intron and spliced
coverage. Finally, intergenic:intergenic showed only a small fraction of base pairs with 
any of either coverage or spliced coverage. In all conflicts the amount of RNAseq support
fell between the cases where tools agreed, indicating that all options were at least
capable of finding weak-spots in the other annotations.
In cases of conflict, Helixer's predictions were substantially more consistent with
the RNAseq data than those of Augustus. Specifically, for base pairs where one tool said CDS and
the other intergenic or intron \ref{}, there was more coverage when 
Helixer said CDS. Similarly, for base pairs where one tool said intron and the other 
CDS or intergenic, there was more spliced coverage when Helixer said intron \ref{}. 
These patterns were consistent in direction but varied in magnitude in the individual species,
with the expception of {\it P. marinus} where Helixer and Augustus performed comparably
at differentiating CDS and introns.

RNAseq based comparison of Helixer and the reference was less clear cut.
Averaged across species, helixer's predictions received slightly more support
than the reference when differentiating CDS, UTR and introns from intergenic;
however helixer's predictions received slightly less support than the reference 
when differentiating CDS and UTR from introns. Performance varies between individual
species, from {\it P. marinus}, where the reference receives more support
in every conflict, to {\it T. cacao}, where Helixer models receive equivalent
or more support in every conflict. Interestingly, these are the two species that
were selected because they had (relative to the usual difference) low Subgenic F1 compared to 
Augustus; for the former the RNAseq confirms relatively weak performance for Helixer, while the 
for the latter RNAseq rather indicates a sub-par reference.

\section{Discussion}
We indtroduce Helixer a novel, deep-learning based tool for gene annotation
prediction. Helixer outperforms AUGUSTUS on base pair wise metrics and on 
consistency with independent RNASeq data.

We include trained models for plants and animals which achieve 
high prediction accuracy on cross-species gene annotation for broad 
phylogenetic groups (land plants and vertebrates, respectively).
Within these groups, this eliminates the dependency on retraining and the
expertise and data required therefore. Production of comparable, single-method
annotations for broad groups has the potential to greatly facilitate downstream
analyses which suffer when data is heterogeneous \citep{}.

We found our models to be highly sensitive to the training genomes we chose. A 
different set could lead to a significant shift in strengths and weaknesses of the 
model and a larger and more spread out set of high quality genomes could also result 
in a wider range of genomes with decent predictions. For this, more computational 
resources would be required. Simply training with a similar amount of data, 
but for a different group (e.g. invertebrates of fungi) could also be used to 
increase the functional predictive range.

@@

Finally, developent of a post-processing method
to go from base pair wise predictions to integrated predictions for whole 
transcripts at each loci could both further improve performance and would
greatly increase real world applications.

@@

HHU HPC-team, deNBI, Ebenhöh lab for computational (particularly GPU) resources.
